---
title: "Celda module analysis tutorial"
author: "Salam AlAbdullatif"
date: "2024-07-19"
output: html_document
---

```{r setup, include=FALSE}
library(dplyr)
library(Seurat)
library(patchwork)
library(celda)
library(singleCellTK)
library(knitr)
library(kableExtra)
library(ggplot2)
knitr::opts_chunk$set(echo = TRUE)
```

## Create Seurat Object

To begin, create a Seurat object for your data and generate cluster labels for your data. In this example, we will be using a PBMC dataset. The steps below outline a basic Seurat analysis. You may also import a Seurat object if one has already been generated, then use the `convertSeuratToSCE` function to proceed. 

```{r, warning=FALSE}
#Seurat pipeline:
##Read in 10X output
pbmc.data <- Read10X(data.dir = "./filtered_gene_bc_matrices/hg19/")
##Create Seurat object
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)
##Normalize data
pbmc <- NormalizeData(pbmc, verbose = FALSE)
##Feature selection
pbmc <- FindVariableFeatures(pbmc, verbose = FALSE)
##Linear transformation
pbmc <- ScaleData(pbmc, verbose = FALSE)
pbmc <- RunPCA(pbmc, verbose = FALSE)
pbmc <- RunUMAP(pbmc, dims = 1:10, verbose = FALSE)
##Generate clusters
pbmc <- FindNeighbors(pbmc, dims = 1:10, verbose = FALSE)
pbmc <- FindClusters(pbmc, resolution = 0.5, verbose = FALSE)
```


### UMAP

```{r, warning=FALSE}
DimPlot(pbmc, reduction = "umap", label = TRUE)
```


```{r, warning=FALSE}
#Convert the Seurat object to a SingleCellExperiment object 
pbmcSce <- convertSeuratToSCE(pbmc, normAssayName = "logcounts")
```


## Generating feature modules from pre-determined clusters

Generate Celda feature modules based on cell clusters that you have provided through the `recursiveSplitModule` function:

```{r}
#Convert Seurat cluster IDs, as Celda requires clusters be numeric vectors starting from 1 (Seurat cluster 0 = Celda cluster 1)
pbmcSce$seurat_clusters <- as.numeric(as.character(pbmcSce$seurat_clusters)) + 1

pbmcSce <- selectFeatures(pbmcSce)
#Run recursiveSplitModule; Seurat clusters will be imported in the zInit parameter.
rsmRes <- recursiveSplitModule(x = pbmcSce, useAssay = "counts", altExpName = "featureSubset", initialL = 2, maxL = 30, zInit = pbmcSce$seurat_clusters, verbose = FALSE)
```

```{r}
#Identify the optimal L (number of modules) using the elbow plot
plotRPC(rsmRes)
```

```{r}
#Subset Celda object
sce <- subsetCeldaList(rsmRes, list(L = 15))
```

```{r}
featureTable <- featureModuleTable(sce, useAssay = "counts", altExpName = "featureSubset")

kable(featureTable, style = "html", row.names = FALSE) %>%
                 kable_styling(bootstrap_options = "striped") %>%
                 scroll_box(width = "100%", height = "500px")
```

For instance, we can see that each module represents a set of co-expressing features, such as module L7 (B lymphocytes - CD79A, MS4A1, TCL1A) and module L8 (NK cells - NKG7, GNLY)

## Run DE on Modules

Differential expression tools can also be used on the gene modules in order to identify the mdoules that define each cell cluster. To do this, first run the `factorizeMatrix` function, which will generate a matrix that measures the contribution of each gene module to each cell population.

```{r DE modules, message=FALSE}
factorize <- factorizeMatrix(sce, useAssay = "counts", altExpName = "featureSubset")
### Take the module counts, log-normalize, then use for DE, violin plots, etc
factorizedCounts <- factorize$counts$cell
factorizedLogcounts <- normalizeCounts(factorizedCounts)
reducedDim(sce, "factorizeMatrix") <- t(factorizedLogcounts)
sce <- runFindMarker(sce, useReducedDim = "factorizeMatrix", cluster = "seurat_clusters", useAssay = NULL)

kable(sce@metadata$findMarker, style = "html", row.names = FALSE) %>%
                 kable_styling(bootstrap_options = "striped") %>%
                 scroll_box(width = "100%", height = "500px")
```


## Generate a Decision Tree

To identify which modules distinguish which clusters, we can use use decision trees implemented in the findMarkersTree function. Assuming that celda has been run on the SingleCellExperiment object, we first get the factorized counts for the modules (modules x cells) and the cluster labels, and input to the findMarkersTree function. 

```{r, message=FALSE}
classes <- as.integer(celdaClusters(sce)) # needs to be an integer vector 
DecTree <- findMarkersTree(factorizedCounts, classes)
```

The output decision tree can be visualized through a dendrogram using the `plotDendro` function:

```{r, warning=FALSE}
plotDendro(DecTree)
```
